Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography

### 2023-09-14 manuscript/wp4.tex
ms 14/09

In this paper, we simulate the processes of differentiation and conservation experimentally.

@@

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}. To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission~+ Communication.

@@

The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, just as in Fig.~\ref{stimuli}. For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension. We describe such a system as ``redundant'' because shape is always expressed by the stem, so the suffix spellings that emerged in this case therefore convey no additional information. Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}. Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O). In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6). Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable. In particular, there are two main kinds of color-expressive system that emerge. The first are the generalization-based expressive systems: K5, M1, P5, and S2. In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape. For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>. In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->). We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner, resulting in low communicative accuracy, and not sustained or elaborated~on in subsequent generations.

@@

The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4. In M7, for example, the participant added <r>, <g>, and <b> (presumably red, gold, blue) to the ends of the words, although the participant's partner only reciprocated the <g> spelling and seemingly failed to understand what was meant by <r> and <b>. In the case of Q7, the pair of participants added <r>, <o>, and <b> (presumably red, orange, blue) to communicate with high accuracy, but, although this system was retained into Generation~8, it started to disintegrate by Generation~9. Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that express both shape and color simultaneously. Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color. In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-pink> or <-red>, <-oran> or <-yell>, and <-blue>), but these pairs were excluded and replaced before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words. We allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red). Indeed, as we see in the results, subsequent generations often did not pick~up~on the color letters (L9, M8, and Q9).} We mention these cases, however, because they are still examples of a conscious effort to differentiate---something that never happened in the Transmission-only condition.

@@

To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate. These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}. The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing. To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content. To do this, we formalized the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another. Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content. The expressive and redundant systems are also considered quite dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we computed a probability distribution over the typological space according to the respective distances from the four reference systems. This approach gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate). The typological probability distribution $t$ over the set of four reference systems $R$ was calculated for participant system $s$ according to% $t(i) \propto e^{-2\mathrm{VoI}(R_i, s)^2}$,
\begin{equation}
	t(i) \propto e^{-2 \cdot \mathrm{VoI}(R_i, s)^2}
\end{equation}
and averaged over chains for a given generation.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
	\vspace*{2pt}
	\caption{Typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). At Generation~0, all 10 chains in both conditions use a holistic system (by design). In Transmission-only, the systems switch first to a redundant and then to a degenerate patterning. In Transmission + Communication, the systems ultimately become degenerate, but also tend to favor holistic systems over redundant ones along the way. Expressive systems were rare in both conditions.}
	\label{typ_dist_dif}
	\end{figure}

The results are plotted in Fig.~\ref{typ_dist_dif}. These probability distributions can be read as the proportion of chains that adopted each of the four configurations at each generation. At Generation~0, all chains are holistic by design. In Transmission-only, there is an initial rapid increase in the number of chains classified into the redundant typological category. The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy. In Transmission + Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common. There is, however, a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings.

As a quantitative measure of how informative the written language is over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}, which expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. Formally, it is given by% $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
	\begin{equation}
	\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
	\end{equation}
where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_dif.eps}}
	\vspace*{2pt}
	\caption{Communicative cost by generation under Transmission-only (dark green; left) and Transmission + Communication (orange; right). The thin lines show each individual transmission chain (with a small amount of jitter added). The thick lines and shaded areas show the mean and 95\% HDI of a regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. The black dotted lines show the communicative cost of the spoken language for comparison. In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure.}
	\label{cost_dif}
	\end{figure}

Fig.~\ref{cost_dif} plots communicative cost over the nine generations of cultural transmission in each condition. In the upper panels, each thin line represents an individual transmission chain, and the thick lines show the mean across chains. The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling. In both conditions, cost increases with generation, implying that the languages become less informative over time. This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent. In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems. This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization. Interestingly, however, cost plateaus at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/posterior_dif.eps}}
	\vspace*{2pt}
	\caption{Posterior parameter estimates from the Experiment~1 regression model. In the upper panels, the dark green curves show results for the Transmission-only condition and the yellow curves show results for the Transmission + Communication condition. The dotted line in the $\alpha$ panel shows the cost of the spoken language. The lower panels show the posterior differences between conditions along with the 95\% HDI credible interval.}
	\label{posterior_dif}
	\end{figure}

To evaluate this argument statistically, we regressed cost on generation in the following Bayesian third-degree polynomial regression model:
	\begin{align*}
	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
	 \mu_i & =    \alpha + \beta_1 g_i + \beta_2 g_i^2 + \beta_3 g_i^3 \\
	\sigma & \sim \mathrm{Exponential}(1)
	\end{align*}
where $g$ is a centered generation number (i.e., Generations~1 through 9 are represented as~-4 through +4). The model states that the cost~$c$ of participant~$i$'s output system is drawn from a normal distribution, the mean of which is determined by a third-degree polynomial model. The intercept $\alpha$ corresponds to Generation~5 (the central generation), and the $\beta$ parameters determine the shape of the polynomial. $\sigma$ represents an across-participant noise term and is given a regularizing exponential prior. The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors. For the full model specification, see the Supplementary Material.

The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}. Essentially, the $\beta$ parameters do not differ substantially between conditions, but the $\alpha$ parameters do appear to differ. This may be interpreted as the polynomial regression models adopting the same shape but plateauing at different levels. Indeed, the posterior difference in $\alpha$, and its credible interval, is less than zero, pointing to an effect of condition on the level at which cost plateaus.

To give greater meaning to these parameter estimates, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$. If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-1.298} = 40.7$\% in Transmission-only and $2^{-0.985} = 50.5$\% in Transmission + Communication, representing a 9.8 percentage point advantage for the communicative condition. However, while the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue). However, such differentiation was not very forthcoming, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure.

So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color. Faced with variation that could be conditioned on either dimension, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion}, \textit{transmission}, and \textit{invention}).

It is particularly surprising that participants chose not to innovate novel systems in communication, despite their being a clear financial incentive to do so. Why might this be the case? Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training. Participants who adopted this interpretation may have feared diverging from what they learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules. Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners? They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.

These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space. Deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>). Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the distinction? Differentiation is hard in the written modality.

@@

The methods were identical to Experiment~1 with two main exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly uninformative over time.

@@

Unlike Experiment~1, no participant pairs attempted to use English color letters and there was only one case of a pair using English color words in Generation~9 of Chain~K, although this pair was excluded and replaced (this generation was the only case in which the training input was fully degenerate, which would have resulted in a strong pressure to find a communicative solution in the form of English). Presumably, the general conservation of expressive spelling in Experiment~2 negated the need to innovate novel systems.

>>>>

In this paper, we simulate the processes of differentiation and conservation experimentally using the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. Following several \textit{generations} of this process, the linguistic system can gradually adapt to the learning biases of the human learners, yielding interesting emergent phenomena, such as compositionality \parencite{Kirby:2008, Kirby:2015}, combinatoriality \parencite{Verhoef:2015}, category structure \parencite{Carr:2017, Carr:2020}, and regularization \parencite{Smith:2010, Ferdinand:2019}. Iterated learning has also been explored in a wide range of domains, such as gesture \parencite{}, population structure \parencite{}, and blah \parencite{}. For reviews, see \textcite{}.

@@

To simulate the cultural evolution of orthography, we adopt the experimental iterated learning paradigm \parencite{Kirby:2008, Kirby:2015}. In this paradigm, some linguistic system is passed along a \textit{transmission chain} of human participants, resulting in repeated cycles of learning and production. Participant $i$ in a transmission chain learns the system based on the linguistic output of participant $i-1$ and then produces new linguistic output for participant $i+1$ to learn from. To explore the hypotheses outlined above, we conducted the experiment under two conditions: Transmission-only and Transmission~+ Communication.

[make this more specific to the present experiment - what is passed down the chain - preview the methods coming up]

@@

The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, following the same 3×3 layout used in Fig.~\ref{stimuli}. The color-coding of these matrices indicates similarity in suffix form: Similar colors are used to represent similar suffixes, making it easier to see how the suffixes pattern with meaning.\footnote{The color-coding was generated independently for each chain by selecting $n$ evenly-spaced hues, where $n$ is the number of unique forms that emerged across the chain, and mapping these hues onto the suffixes in alphabetical order.} For example, the system at Generation~9 in Chain~D uses three spellings (<koe>, <ko>, and <co>) to express the shape dimension, yielding a horizontal stripes pattern in the matrix representation. We describe such a system as ``redundant'' because shape was consistently and reliably expressed by the stem, so the suffix spelling system that emerged in this case conveys no additional information---the suffix simple repeats whichever shape was marked by the stem. Redundant suffix systems are characteristic of the Transmission-only condition, with similar outcomes occurring in Chains~C, E, F, and~H. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

@@

The results for the Transmission~+ Communication condition (Chains~K--T) are shown in Fig.~\ref{dif_com}. Like Transmission-only, degeneration to a single spelling by the ninth generation is a relatively common outcome (e.g., Chains~N, R, and~S and to a lesser extent Chains~L, M, and~O). In contrast, redundant, shape-expressive systems (as indicated by horizontal stripes) are relatively rare (e.g., K8, M6 and Q1--6). Instead, the presence of the communicative task appears to favor color-expressive systems, although these are far from common and often unstable. In particular, there are two main kinds of color-expressive system that emerge. The first are the generalization-based expressive systems: K5, M1, P5, and S2. In these cases, the participant tended to generalize their input in a way that is consistent with expressing color over shape. For example, in K5, pink was consistently spelled <ko> and blue was consistently spelled <co>. In the case of M1, color is expressed by the final vowel letters (<oe> for pink, <o> for yellow, and <oh> for blue) with the spelling of the /k/ sound conditioned on the stem (<buvic->, <zetik->, and <wopiq->). We note, however, that in all these cases the expressive system was not reciprocated by the participant's partner---resulting in low communicative accuracy---and not sustained or elaborated~on in subsequent generations.

The second type of color-expressive system is one that simply appropriates the expressive power of English to differentiate color: K9, M7, Q7, and to a lesser extent L7 and O4. In M7, for example, the participant added <r>, <g>, and <b> (presumably red, gold, blue) to the ends of the words, although the participant's partner only reciprocated the <g> spelling and seemingly failed to understand what was meant by <r> and <b>. In the case of Q7, the pair of participants added <r>, <o>, and <b> (presumably red, orange, blue) to communicate with high accuracy, but, although this system was retained into Generation~8, it started to disintegrate by Generation~9. Overall, in the five cases where English color letters were used, the systems did not really catch~on, perhaps because they retained spelling redundancy from the previous generation, resulting in unnecessarily complex suffix spellings that were too difficult to learn for subsequent generations. Q7, for example, uses <-coe->, <-co->, <-qo-> for shape plus <-r>, <-o>, <-b> for color. In addition to this handful of cases, there were a further four pairs of participants who used full English words as the suffix (typically <-pink> or <-red>, <-oran> or <-yell>, and <-blue>), but these pairs were excluded and replaced before iteration to the next generation.\footnote{In our preregistration, we stated that we would reject participants who used English color words. We allowed single letters because these were typically less transparent, especially in cases where the choice of letter (e.g., <g> or <r>) did not match how another person might have perceived the color (yellow vs.\@ gold vs.\@ orange or pink vs.\@ red). Indeed, as we see in the results, subsequent generations often did not pick~up~on the color letters (L9, M8, and Q9).} We mention these cases, however, because they are still examples of a conscious effort to differentiate---something that never happened in the Transmission-only condition.

@@

To analyze more formally what types of system tend to emerge under different conditions, we first designated four suffix systems that are of particular interest: holistic, expressive, redundant, and degenerate. These four typological categories may be positioned along the simplicity--informativeness continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}. The holistic\footnote{Our use of the term \textit{holistic} is slightly unusual here, owing to the fact that we are focused on the suffix level and not the word level. By holistic, we only mean that each shape--color combination has a unique suffix form, so the nine suffixes must all be learned as separate cases. We do not distinguish between truly holistic suffixes (nine unique suffixes with no structure in how they relate to each other) and compositional suffixes (nine unique suffixes that can be generated from compositional rules, as was the case in, for example, Q8).} system has nine unique forms each of which expresses a particular color--shape combination. This is complex to learn, but the suffix alone can pick out exactly one meaning. The expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant). These systems are easier to learn, but only the expressive system is fully informative (when acting in combination with the stem). The degenerate system uses just one suffix form. This makes it trivial to learn, but entirely uninformative. To classify a participant's output into one of these four typological categories, we computed which of the reference systems was most similar in terms of its information content. To do this, we formalized the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite{Meila:2007} defined as the distance metric between any two such partitions.\footnote{Variation of information is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one partition into another. Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content. The expressive and redundant systems are also considered quite dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} A given participant's output system is then classified into one of the four typological categories based on whichever reference system is closest.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/typ_dist_dif.eps}}
	\vspace*{2pt}
	\caption{Upper panels show the typological distribution at each generation and in each condition across four typological categories: holistic (H; purple), expressive (E; blue), redundant (R, green), and degenerate (D; yellow). Bottom panels show posterior differences (mean and 90\% HDI) in the proportion of chains falling into each typological category; positive values signify greater use of a given category in the communicative condition.}
	\label{typ_dist_dif}
	\end{figure}

The typological distributions are plotted in Fig.~\ref{typ_dist_dif}, reveling the proportion of the 10 chains that fall into each typological category at each generation. In Transmission-only, the holistic systems used to initialize the chains are rapidly replaced by redundant systems by Generation~2. The dominance of the redundant category is then gradually eroded as the chains transition to degeneracy. In Transmission~+ Communication, there is initially a fairly even mix of all four kinds of system, but by Generation~9, degenerate systems tend to be most common. There is also a notable increase in holistic systems emerging in later generations; these are the cases of compositional suffixes that arose through the addition of English color letters on top of redundant suffix spellings (i.e., K9, M1, M7, and Q7). For a statistical comparison across conditions, we adopt a Dirichlet-Multinomial model of typological distribution $t$ for generation $i$,
	\begin{equation}
	\begin{aligned}
	   t_i & \sim \mathrm{Multinomial}(10, \theta_i) \\
	   \theta_i & \sim \mathrm{Dirichlet}([1, 1, 1, 1]),
	\end{aligned}
	\end{equation}
and compute the posterior difference in $\theta_i$ (the estimated proportion of chains falling into each category) across conditions: $\Delta(\theta_i) = \theta_{i_\mathrm{comm}} - \theta_{i_\mathrm{trans}}$. The means and 90\% HDIs of these posterior differences are plotted in the bottom panels of Fig.~\ref{typ_dist_dif}. In general, we find no clear difference in the number of expressive systems between the two conditions; however, we do find that redundant systems are less common in Transmission~+ Communication (in most generations), while holistic systems tend to be a little more common (most notably at Generation~7). Although there is not much evidence to suggest that Transmission~+ Communication strongly favors more informative systems (i.e., the expressive and holistic categories), it is interesting to note that the communicative condition did seem to disfavor the inexpressive redundant systems.

As a more general measure of how informative the written words are over time (i.e., when the stem and suffix act in combination), we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}. This measure expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. It is given by
	\begin{equation}
	\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
	\end{equation}
where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_dif.eps}}
	\vspace*{2pt}
	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange). The thin lines show each individual transmission chain (with a small amount of added jitter). The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. The black dotted lines show the communicative cost of the spoken language for comparison. In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure. \textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
	\label{cost_dif}
	\end{figure}

Fig.~\ref{cost_dif}A plots communicative cost over the nine generations of cultural transmission in each condition. The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling. In both conditions, cost increases with generation, implying that the languages become less informative over time. This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent. In contrast, this is not the expected result in Transmission~+ Communication, where the communicative pressure should promote the emergence of informative systems. This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization. Interestingly, however, cost plateaus at a slightly lower level in Transmission~+ Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.

To evaluate this argument statistically, we fit the following regression model:
	\begin{equation}
	\begin{aligned}
	   c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
	 \mu_i & =    \alpha + \beta_1 g_i + \beta_2 g_i^2 + \beta_3 g_i^3 \\
	\sigma & \sim \mathrm{Exponential}(1),
	\end{aligned}
	\end{equation}
where $g$ is a centered generation number (i.e., Generations~1 through 9 are represented as~$-4$ through $+4$). The model states that the cost~$c$ of participant~$i$'s output system is drawn from a normal distribution, the mean of which is determined by a third-degree polynomial model. The intercept $\alpha$ corresponds to Generation~5 (the central generation), and the $\beta$ parameters determine the shape of the polynomial. $\sigma$ represents an across-participant noise term and is given a regularizing exponential prior. The priors on $\alpha$ and $\beta$ are not shown here, for brevity, but these parameters are estimated separately for each chain and pooled into global estimates with the use of Normal hyperpriors. For the full model specification, see the Supplementary Material. The posterior parameter estimates of the model are shown in Fig.~\ref{cost_dif}B. Essentially, the $\beta$ parameters do not differ substantially between conditions, but the $\alpha$ parameters do differ. This may be interpreted as the polynomial regression models adopting approximately the same shape but plateauing at different levels of cost. Additionally, we find that the $\alpha$ parameters themselves are lower than the cost of the spoken language (indicated by the dotted line), although it is unclear how this situation might unfold beyond Generation~9, especially in Transmission-only where the model predicts that cost will continue to increase further until all orthographic systems become aligned with the spoken language.

Finally, cost may be converted into the expected probability of communicative success with the equation $p=2^{-c}$. If we take Generation~5 as representative of the level at which cost plateaus, we find that estimated communicative success is $2^{-\alpha} = 40.8$\% (90\% HDI: 35.1--46.3) in Transmission-only and 50.6\% (90\% HDI: 45.9--54.9) in Transmission~+ Communication, representing an advantage of around 10 percentage points for the communicative condition. Overall, then, while the presence of the communicative pressure does appear to make a meaningful difference to the informativeness of the emergent systems, the levels of informativeness that were achieved were not especially impressive considering that an expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use, while also being fairly easy to learn.

@@

Our first experiment tested whether informative, heterographic orthography could emerge through differentiation and whether it would emerge preferentially under communicative pressure. Overall, there was little evidence of systematic orthographic differentiation. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue). Such differentiation was, however, not very forthcoming, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was usually conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure \parencite[e.g.,][]{Kirby:2015}.

So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. Faced with variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expulsion}, \textit{transmission}, and \textit{station}). As we will see in the following experiment, it is indeed possible for participants to learn expressive systems, even under full homophony; the lack of such outcomes in the present experiment suggests a reluctance to defy the spoken language.

It is particularly surprising that participants chose not to innovate novel systems in the communicative condition, despite their being a clear financial incentive to do so. Why might this be the case? Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training. Participants who adopted this interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules. Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners? They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems. These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space. Deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>). Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the distinction? Differentiation is hard in the written modality.

@@

The methods were identical to Experiment~1 with two exceptions: The artificial language used to seed each chain started out fully compositional (in both its spoken and written forms), and two sound mergers were artificially induced during cultural transmission, resulting in the spoken forms of the suffixes becoming increasingly homophonous and uninformative over time.

@@

Unlike Experiment~1, no participant pairs attempted to use English color letters and there was only one case of a pair using English color words (Generation~9 of Chain~K), although this pair was excluded and replaced (this generation was the only case in which the training input was fully degenerate, which would have resulted in a strong pressure to find a communicative solution in the form of English). Presumably, the general conservation of expressive spelling in Experiment~2 negated the need to innovate novel systems.
